We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-01-31 paper/paper.tex
edit

\begin{abstract}
We use simulations to build an explicit probability distribution for the
asphericity in the satellite distribution around galaxies similar to the
Local Group (LG) in the Lambda Cold Dark Matter (LCDM) paradigm.
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-Dark and ELVIS. 
Detailed results differ greatly among the simulations, suggesting a
strong influence of resolution, simulated physics and criteria to select
LG pairs.
However, there are three common trends in these simulations. First, at most of
$1\%$ the pairs are expected to have satellite distributions with the
same asphericity as the Local Group; second, between
$38\%$ to $59\%$ of the pairs have a halo with a satellite
distribution as aspherical as in M31; and third, at most $2\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The brightest satellites in M31 galaxy turn out to have a rather typical distribution 
in the LCDM context, while the MW is atypical in this respect. 
These results place the LG close to the level of a 3-$\sigma$ outlier
in the LCDM paradigm.
The approach presented here will facilitate the comparison between
asphericity results from different numerical setups and choices to
select halo pairs in simulations, hopefully helping to clarify the
reasons behin the LG atipicallity in LCDM. 
\end{abstract}

@@

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{gaussian_model_illustrisdm_M31.pdf}
\includegraphics[width=0.45\textwidth]{gaussian_model_illustrisdm_MW.pdf}
\caption{Correlations between the multivariate gaussian model built on
  the normalized values for the plane width $w$, $c/a$ ratio and $b/a$ ratio. 
Left/right panel correspond to M31/MW. 
The contour levels in the 2D histograms correspond to the $1\sigma$,
$2\sigma$ and $3\sigma$ contours in two dimensions. 
The dahsed vertical lines in the histograms along the diagonal
correspond to the $1\sigma$ boundaries in one dimension.
The results for the gaussian model are built from $10^6$ point
realizations in the six-dimensional space spaned by the variables of
interest. 
The correlation matrix and the mean values are computed from the
results in the Illustris-1 simulation.
The cross indicates the LG values.
This plot clearly shows how the M31 results are well within the
expectations from simulations while MW has an unusual low value for
the plane width and the $c/a$ axis ratio.
Equivalent results from the ELVIS data are presented in Figure
\ref{fig:correlations_elvis}.
\label{fig:correlations_illustris}}
\end{figure*}

@@

Figure \ref{fig:scatter_width} summarizes the results for the width
measurements.
The left panel compares the results for the MW and M31
observations (stars) against its spherically randomized satellites.
The most interesting outcome is that the MW plane width is smaller
than $\approx 98\%$ of the planes computed from the randomized distribution,
while the M31 plane width is only slightly smaller than the average of
the distribution.

@@

\subsection{$c/a$ axis ratio}
Figure \ref{fig:scatter_ca_ratio} shows the results for the minor to
major axis ratio. 
Corresponding in the ELVIS data are in the appendix in Figure
\ref{fig:scatter_elvis}.

The left panel shows the results for the LG compared against its
spherically randomized version.
As expected the $c/a$ ratio in the MW is significantly lower as the
measured values for spherical distributions. 
On the other hand the ratio for M31 is lower than the mean of the
spherical values but still well within its variance.
In contrast to the results for the plane width, this time the
randomized values have symmetrical expectations between the two
galaxies.

The middle panel shows the LG compared against the results from
Illustris. In this case we find a similar trend. The MW is atypical
and M31 is within the variance from the simulation data.
This time, however, there is a single MW-like galaxy out of the total
of 20 that shows an $c/a$ as small as the MW.
The only assymetry evident between the simulated halos is in the
dispersion, the $c/a$ distribution for the massive halos seems to be
wider than it is for the less massive partner.

The right panel shows the normalized results. 
This highlights the two results mentioned above.
First, the MW shows a low $c/a$ ration close to two standard deviations away
from the mean value of the spherical distribution; this contrasts the
results for M31 which are close to $1$ standard deviation away.
Second, the less massive halos in the Illustris simulation present a
smaller dispersion in their deviations from sphericity than the
its masssive partner.
The MW axis ratio smaller than $\approx 98\%$ of the randomized
satellite distributions.   and its atipicallity is only comparable to
one halo in the Illustris  simulation (none in the ELVIS simulation).

@@

Figure \ref{fig:scatter_ba_ratio} shows the results for the minor to
major axis ratio. 
Corresponding in the ELVIS data are in the appendix in Figure
\ref{fig:scatter_elvis}.
In all cases of comparison (against randomized distribution,
comparasion against simulations and degree of deviation from
sphericity) the results for both the MW and M31 do not seem atypical
or deviated from the expectations for an spherical dsitribution.

>>>>

\begin{abstract}
We use numerical simulations to build an explicit probability distribution for the
asphericity in the satellite distribution around galaxies similar to the
Local Group (LG) in the Lambda Cold Dark Matter (LCDM) paradigm.
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-Dark and ELVIS. 
Detailed results differ greatly among the simulations, suggesting a
strong influence of resolution, simulated physics and criteria to select
LG pairs.
However, there are three common trends in these simulations. First, at most of
$1\%$ the pairs are expected to have satellite distributions with the
same asphericity as the Local Group; second, between
$38\%$ to $59\%$ of the pairs have a halo with a satellite
distribution as aspherical as in M31; and third, at most $2\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The brightest satellites in M31 galaxy turn out to have a rather typical distribution 
in the LCDM context, while the MW is atypical in this respect. 
These results place the LG close to the level of a 3-$\sigma$ outlier
in the LCDM paradigm.
The approach presented here will facilitate the comparison between
asphericity results from different numerical setups and choices to
select halo pairs in simulations, hopefully helping to clarify the
reasons behin the LG atipicallity in LCDM. 
\end{abstract}

@@

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{gaussian_model_illustrisdm_M31.pdf}
\includegraphics[width=0.45\textwidth]{gaussian_model_illustrisdm_MW.pdf}
\caption{Results from multivariate gaussian model fitted on   the
  normalized values for the plane width $w$, $c/a$ ratio and $b/a$
  ratio from the Illustris1-Dark results.  
  Left/right panel correspond
  to M31/MW.  
  The cross indicates the observed values.
  The contour levels in the 2D histograms correspond to the $1\sigma$,
$2\sigma$ and $3\sigma$ contours in two dimensions.
  The dahsed vertical lines in the histograms along the diagonal
  correspond to the $1\sigma$ boundaries in one dimension.
  The results for the gaussian model are built from $10^6$ point
  realizations in the three-dimensional space spaned by the variables of
  interest for each halo. 
  This plot clearly shows how the M31 results are well within the
  expectations from simulations while the MW has an unusual low value for
  the plane width and the $c/a$ axis ratio.
\label{fig:correlations_illustrisdm}}
\end{figure*}

@@

Figure \ref{fig:scatter_width} summarizes the results for the width
measurements.
The left panel compares the results for the MW and M31
observations (stars) against its spherically randomized satellites
(circles). 
The most interesting outcome is that the MW plane width is smaller
than $\approx 98\%$ of the planes computed from the randomized distribution,
while the M31 plane width is only slightly smaller than the average of
the distribution.

@@

\subsection{$c/a$ axis ratio}

Figure \ref{fig:scatter_ca_ratio} shows the results for the minor to
major axis ratio. 
The layout is the same as in Figure \ref{fig:scatter_width}.
The results for the $c/a$ ratio follow the same trends as for the
width $w$.

The left panel in Figure \ref{fig:scatter_ca_ratio} shows how the MW
the $c/a$ ratio is significantly lower as the measured values for
spherical distributions, it is smaller than $\approx 98\%$ of the
randomized distributions.   
On the other hand the ratio for M31 is lower than the mean of the
spherical values but still well within its variance.

The middle panel in the same Figure shows the LG compared against the
results in the simulations. 
In this case we find a similar trend as before. 
The MW is atypical and M31 is within the variance from the simulation data.
This time, however, there are two MW-like halos out of the total
of 24 that show an $c/a$ as small as the MW.

The right panel shows the normalized results. 
The MW shows a low $c/a$ ration close to between two and three
standard deviations away from the mean value of the spherical
distribution; this contrasts the results for M31 which are close to
$1$ standard deviation away.

@@

Figure \ref{fig:scatter_ba_ratio} shows the results for the minor to
major axis ratio with the same layout as Figure \ref{fig_scatter_ca_ratio}
In all cases of comparison (against randomized distribution
and simulations  the results for both the MW and M31 are typical.
